Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-04 _tex/index.tex
Built site for gh-pages

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(mean), referred to as the \emph{modal discriminal process}, indicates
the stimulus' position on the trait continuum, and its dispersion, known
as the \emph{discriminal dispersion}, reflects the variability in the
perceived trait level of the stimulus.

@@

On the one hand, the assumption of \emph{equal dispersion between
stimuli} suggests that variability in perceived trait levels is
consistent across all stimuli. However, Thurstone observed that this
assumption may not hold when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits
and stimuli can introduce greater judgment discrepancies due to their
unique characteristics. In fact, research on text quality may already
provide evidence of the violation of this assumption, particularly
through the use of misfit statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
Misfit statistics quantify the judgment discrepancies associated with a
given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
In this context, texts identified as ``misfits'' display greater
judgment discrepancies than other texts. This implies that the
discriminal difference associated with these texts shows a broader
dispersion, suggesting that their discriminal processes also exhibit
greater variation than other texts. This reasoning also applies to
``misfit'' judges, whose evaluations reflect substantial deviations due
to the unique or extreme characteristics of the stimuli.

{what are the challenges when the equal dispersion assumption is
violated?}

>>>>

In its most general form, Thurstone's theory deals with pairwise
comparisons of stimuli made by a single judge
\citep[pp.~267]{Thurstone_1927b}. The theory proposes that two key
factors determine the dichotomous outcome of these comparisons: the
discriminal process of each stimulus and their discriminal difference.
The \emph{discriminal process} represents the psychological impact each
stimulus has on judges or, more simply, their underlying perception of
the stimulus' trait level. According to the theory, the discriminal
process for each stimulus follows a Normal distribution, where its mode
(mean), referred to as the \emph{modal discriminal process}, indicates
the stimulus' position on the trait continuum, and its dispersion, known
as the \emph{discriminal dispersion}, reflects the perceived trait
variability of the stimulus.

@@

On the one hand, the assumption of \emph{equal dispersion between
stimuli} suggests that the perceived trait variability is consistent
across all stimuli. However, Thurstone observed that this assumption may
not hold when researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
introduce judgment discrepancies due to their unique characteristics.
Research on text quality may already provide evidence of this
assumption's violation, particularly when considering the role of misfit
statistics in the analysis of CJ data
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
Misfit statistics quantify the judgment discrepancies associated with a
given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, texts identified as ``misfits'' usually exhibit more
judgment discrepancies than others. This, in turn, implies that the
discriminal difference associated with these texts shows a broader
dispersion, ultimately suggesting that their discriminal processes also
exhibit more variation than other texts. Notably, this reasoning also
applies to ``misfit'' judges, whose evaluations reflect substantial
deviations due to the unique characteristics of the stimuli or the
judges themselves.

Moreover, assuming equal dispersion between stimuli despite its
violation causes Case V (and the BTL model) to overlook critical
differences in the reliability of the trait across stimuli. Moreover, if
researchers recognize that misfit statistics help to identify such
violations, the usual practice in the CJ literature of excluding stimuli
based on these statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases remain unpredictable because they depend on
which stimuli researchers exclude from the analysis. Together, these
oversights undermine the reliability of the trait and ultimately
compromise its validity \citep[pp.~2]{Perron_et_al_2015}.
